dispRity demo

### 2015-10-08 doc/dispRity-tutorial.Rnw
Updated demos

However, I greatly encourage you to use your own data (after all, that's the point of an \R package).
You will simply need (1) an ordinated distance matrix: make sure that the number of column is at maximum the number of rows minus one.
And secondly (2) a phylogenetic tree with the tips matching the ordinated matrix and a \texttt{\$root.time} element for determining the age of the tree.
You can use the function \texttt{tree.age} provided in this package but not described in this demo for generating the root.time automatically (see \texttt{?tree.age}).
In the following examples, just replace all the \texttt{BeckLee} mentions in this demo by your very own data:

@@

<<dispRity class1, eval=FALSE>>=
str(sum_of_ranges)
## That's a 2512 lines output!
@

@@

<<cust.series2>>=
## We want to separate the species around the node 71 (see phylogeny above).
## All the descendant of these node are crown and all the ancestors are stem
crown <- extract.clade(MyData_tree, node = 71)$tip.label
stem <- drop.tip(MyData_tree, tip = crown)$tip.label
## We then have to feed this information in a data frame with one column
factors <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
    ncol = 1, dimnames = list(c(crown, stem))))
## We then can use the customised series function
crown_stem <- cust.series(MyData_mat50, factors)
## This created a dispRity object containing two series: crown and stem
class(crown_stem)
crown_stem
## This object contains three elements
names(crown_stem)
## With "data" being the list sub-matrices
str(crown_stem$data)
## "taxa" being the list of taxa in the original ordinated matrix
str(crown_stem$taxa)
## and finally, "series" containing information on the series type (custom) and names
## (crown and stem)
crown_stem$series
@

Of course, using phylogeny as a factor is just an example.
Any type of factors can be used in this example (e.g. body mass categories, habitat, diet, etc...).

\subsubsection{Time splitting}
\texttt{time.series} is a function that allows to split the matrix in different ways using time as a category.
This functions needs as input an ordinated matrix and a matching phylogenetic tree.
Two types of time series splitting can be performed by using the \texttt{method} option:
\begin{enumerate}
\item the classical discrete time splitting (or time-binning) using \texttt{method = "discrete"};
\item or a new continuous time splitting (or time-slicing) using \texttt{method = "continuous"}.
\end{enumerate}
For the time-slicing method details, see Guillerme and Cooper \textit{in prep.} \cite{GuillermeSTD}.
Once the method is chosen, the function intakes the \texttt{time} argument which can be a vector of \texttt{numeric} values for:
\begin{enumerate}
\item defining the boundaries of the time bins (when \texttt{method = "discrete"});
\item defining the time-slicing places (when using \texttt{method = "continuous"}).
\end{enumerate}
Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a fix number of equidistant time-bins/slices.
Additionally, we can also input a data frame containing the first/last occurrence data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.

@@

We now have three time bins of 44.50368 Ma each. But hey, that's a bit silly, let's stick to bins of 40 Ma.
When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1\% of the the tree length as long as at least three elements are present at that point in time (an extra 1\% is added until reaching the three elements).
We can also ask to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips AND nodes.

@@

<<continuous.series>>=
## Let's generate four time slices every 40 Ma assuming a gradual evolution model
time_slices <- time.series(data = MyData_mat99, tree = MyData_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = MyData_ages)

time_slices
## Note that in the same way as for the discrete method, we can also
## automatically generate the slices
time.series(data = MyData_mat99, tree = MyData_tree, method = "continuous",
    model = "gradual", time = 4)
## But that gives the same silly results (slicing every 44.50368 Ma)
@

@@

This function proposes the user to specify the bootstrap algorithm with the \texttt{boot.type} method.
Currently two algorithms are implemented (but see section \ref{whatsleft}):
\begin{enumerate}
\item \texttt{"full"} where the bootstrapping is entirely stochastic (all the data is bootstrapped)
\item \texttt{"single"} where only one random taxa is replaced by one other random taxa each pseudo-replication
\end{enumerate}
This function also has a \texttt{rarefaction} option to rarefy the data (how unexpected!).
The default status is \texttt{FALSE} but it can be set to \texttt{TRUE} to fully rarefy the data (i.e. remove $n$ taxa for the number of pseudo-replicates, where $n$ varies from the maximum number of taxa present in the dataset to a minimum of 3 taxa).
It can also be set to a fix \texttt{numeric} value (or a set of numeric values).
Finally, the last option from this function, \texttt{rm.last.axis}, is whether to remove a certain amount of axis of the ordinated data.
This option can be logical where \texttt{FALSE} (default) will not remove any axis and \texttt{TRUE} will remove the 5\% last axis or it can be set to a user defined proportion.
This is a common practice in disparity-through-time analysis but it is hard to justify statistically.
Here is a series of examples:

@@

Because disparity can be measured in many ways, this function is a tool to measure disparity \textit{as defined by the user} (and here's where the modulable part comes in).
In fact, the \dispRity function intakes two main arguments: the data and the disparity metric.
The disparity metric boils down to a single value that describes the matrix.
Some are classic like the sum or the product of the ranges or the variances of the matrix columns (i.e. the any-o-space axis/dimensions) \cite{Wills1994}.
However, many more exist like the median distances between the taxa and the any-o-space centroid \cite{GuillermeSTD}.

@@

\subsubsection{\texttt{summary} function}
This function intakes a \dispRity object plus various options namely the \texttt{CI} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and two \textit{visual} options which are whether to recall the \dispRity options and how much digits are wanted in the results.

<<summary>>=
## Let's have a look at the disparity-through-time using in the time-binned data
summary(disp_time_binsBS)
## The CIs are calculated as 50 and 95 and the central tendency is the mean by default.
## We can change that easily by specifying different values in the options
summary(disp_time_binsBS, CI = 25, cent.tend = median, rounding = 0)
## Also, if we happen to have forgotten what was in the disp_time_binsBS object (as in
## which options where used to modify the original matrix) we can use the recall option:
summary(disp_time_binsBS, recall = TRUE)
## Note that the recall just prints text. The output of summary.dispRity is always a
## data.frame
summary_table <- summary(disp_time_binsBS, recall = TRUE)
class(summary_table)

## Finally we can see the results of the rarefaction analysis for the crown/stem data
head(summary(disp_crown_stemBS))
## This outputs a longer table with all the variations of the number of crown/stem taxa
## from the maximum (30 and 20) to the minimum (3)
@

@@

The grey squares represent the confidence intervals (50 in dark grey and 90 in light grey) and the dot represents the mean.
However, it might be interesting to see how many taxa are present in each group.
This can be done via the \texttt{diversity} option.
A third interesting option is to plot the rarefied data if a rarefaction was applied on the data with the \texttt{boot.matrix} function.
For displaying the rarefied data we can use the \texttt{rarefaction} option.
This option can be either \texttt{FALSE} (default) for always plotting the maximum number of taxa per series or \texttt{TRUE} to always plot the minimum.
Additionally, it can be a specific number to plot a specific number of taxa.
Let say we want both groups to have 20 taxa maximum.
We can see now that the confidence intervals overlap more than previously

<<plot1, fig.width=15, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 15, height = 5) ; par(mfrow = (c(1,3)), bty = "n")
## Plotting the score for both groups
plot(disp_crown_stemBS, ylim = c(25,40))
## Same one with the number of taxa for each data set
plot(disp_crown_stemBS, diversity = TRUE, ylim = c(25,40))
## Same one with rarefied data (note how that affects the diversity data as well!)
plot(disp_crown_stemBS, diversity = TRUE, rarefaction = 20, ylim = c(25,40))
@

@@

<<plot2, fig.width=5, fig.height=8, out.width='0.5\\linewidth'>>=
## Graphical options
par(bty = "n")
## The rarefaction curves
plot(disp_crown_stemBS, diversity = TRUE, rarefaction = "plot")
## Note that the function automatically draws the curves for each series and splits
## the plot screen accordingly.
@

@@

Additionally, for the \texttt{type = "discrete"} option, we might also decide to plot many boxplots (which can be a bit messy).
The \texttt{discrete\_type} option allows to switch between \texttt{"box"} or \texttt{"line"} for showing the results in one or the other format.
Let's look at that with the time binned data (even though there is only three time bins).
Finally, we can use the \texttt{type = "continuous"} option that comes in handy for plotting continuous data like in the time sliced analysis (obviously...).

<<plot3, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
## The disparity-through-time data in a time-binned way (with lines rather than
## boxes)
plot(disp_time_binsBS, discrete_type = "line")
## The disparity-through-time data in a time-sliced ways
plot(disp_time_slicesBS)

@@

<<plot5, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Graphical options
par(bty = "n")
## A plot with many options!
plot(disp_time_slicesBS, CI = c(seq(from=10, to=100, by=10)), cent.tend = sd,
    type = "continuous", diversity = "log", col = c("black", rainbow(10)),
    ylab = c("Disparity", "log(Diversity)"), xlab = "Time (in in units from
    past to present)", time.series = FALSE, main = "Many options...")
@

>>>>

However, I greatly encourage you to use your own data (after all, that's the point of an \R package).
You will simply need (1) an ordinated distance matrix: make sure that the number of column is at maximum the number of rows minus one.
And secondly (2) a phylogenetic tree with the tips matching the ordinated matrix and a \texttt{\$root.time} element for determining the age of the tree.
You can use the function \texttt{tree.age} provided in this package but not described in this demo for generating the \texttt{\$root.time} automatically (see \texttt{?tree.age}).
In the following examples, just replace all the \texttt{BeckLee} mentions in this demo by your very own data:

@@

<<dispRity class1, eval=FALSE>>=
str(sum_of_ranges)
## That's a more than 4000 lines of output
@

@@

<<cust.series2>>=
## We want to separate the species around the node 71 (see phylogeny above).
## All the descendant of these node are crown and all the ancestors are stem
crown <- extract.clade(MyData_tree, node = 71)$tip.label
stem <- drop.tip(MyData_tree, tip = crown)$tip.label
## We then have to feed this information in a data frame with one column
factors <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
    ncol = 1, dimnames = list(c(crown, stem), "Group")))
## We then can use the customised series function
crown_stem <- cust.series(MyData_mat50, factors)
## This created a dispRity object containing two series: crown and stem
class(crown_stem)
crown_stem
## This object contains three elements
names(crown_stem)
## With "data" being the list sub-matrices
str(crown_stem$data)
## "taxa" being the list of taxa in the original ordinated matrix
str(crown_stem$taxa)
## and finally, "series" containing information on the series type (custom) and names
## (crown and stem)
crown_stem$series
@

Of course, using phylogeny as a factor is just an example.
Any type of factors can be used in this example (e.g. body mass categories, habitat, diet, etc...).
See the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos} for other specific examples.

\subsubsection{Time splitting}
\texttt{time.series} is a function that allows to split the matrix in different ways using time as a category.
This functions needs as input an ordinated matrix and a matching phylogenetic tree.
Two types of time series splitting can be performed by using the \texttt{method} option:
\begin{enumerate}
\item the classical discrete time splitting (or time-binning) using \texttt{method = "discrete"};
\item or a new continuous time splitting (or time-slicing) using \texttt{method = "continuous"}.
\end{enumerate}
For the time-slicing method details, see Guillerme and Cooper \textit{in prep.} \cite{GuillermeSTD}.
Once the method is chosen, the function intakes the \texttt{time} argument which can be a vector of \texttt{numeric} values for:
\begin{enumerate}
\item defining the boundaries of the time bins (when \texttt{method = "discrete"});
\item defining the time-slicing places (when using \texttt{method = "continuous"}).
\end{enumerate}
Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a fix number of equidistant time-bins/slices.
Additionally, we can also input a data frame containing the First/Last Occurrence Data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.

@@

We now have three time bins of 44.50368 Ma each. But hey, that's a bit silly, let's stick to bins of 40 Ma.
When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1\% of the the tree length as long as at least three elements are present at that point in time (an extra 1\% is added until reaching the required minimum of three elements).
We can also ask to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips \textit{AND} nodes.

@@

<<continuous.series>>=
## Let's generate four time slices every 40 Ma assuming a gradual evolution model
time_slices <- time.series(data = MyData_mat99, tree = MyData_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = MyData_ages)
time_slices
## Note that in the same way as for the discrete method, we can also
## automatically generate the slices
time.series(data = MyData_mat99, tree = MyData_tree, method = "continuous",
    model = "gradual", time = 4)
## But that gives the same silly results (slicing every 44.50368 Ma)
@

@@

This function proposes the user to specify the bootstrap algorithm with the \texttt{boot.type} method.
Currently two algorithms are implemented (but see section \ref{whatsleft}):
\begin{enumerate}
\item \texttt{"full"} where the bootstrapping is entirely stochastic (all the data is bootstrapped)
\item \texttt{"single"} where only one random taxa is replaced by one other random taxa each pseudo-replication
\end{enumerate}
This function also has a \texttt{rarefaction} option to rarefy the data (how unexpected!).
The default status is \texttt{FALSE} but it can be set to \texttt{TRUE} to fully rarefy the data (i.e. remove $n$ taxa for the number of pseudo-replicates, where $n$ varies from the maximum number of taxa present in the dataset to a minimum of 3 elements).
It can also be set to a fix \texttt{numeric} value (or a set of numeric values).
Finally, the last option from this function, \texttt{rm.last.axis}, is whether to remove a certain amount of axis of the ordinated data.
This option can be logical where \texttt{FALSE} (default) will not remove any axis and \texttt{TRUE} will remove the 5\% last axis or it can be set to a user defined proportion.
This is a common practice in disparity-through-time analysis but it is hard to justify statistically.
Here is a series of examples:

@@

Because disparity can be measured in many ways, this function is a tool to measure disparity \textit{as defined by the user} (and here's where the modulable part comes in).
In fact, the \dispRity function intakes two main arguments: the data and the disparity metric.
%To change in future version!
The disparity metric boils down to a single value that describes the matrix.
Some are classic like the sum or the product of the ranges or the variances of the matrix columns (i.e. the any-o-space axis/dimensions) \cite{Wills1994}.
However, many more exist like the median distances between the taxa and the any-o-space centroid \cite{GuillermeSTD}.

@@

\subsubsection{\texttt{summary} function}
This function intakes a \dispRity object plus various options namely the \texttt{quantile} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and two \textit{visual} options which are whether to recall the \dispRity options and how much digits are wanted in the results.

<<summary>>=
## Let's have a look at the disparity-through-time using in the time-binned data
summary(disp_time_binsBS)
## The 50 and 95 quantiles are calculated and the central tendency is the mean by default.
## Note that the function also displays the observed disparity (non-bootstrapped).
## We can change that easily by specifying different values in the options
summary(disp_time_binsBS, quantile = 88, cent.tend = sd, rounding = 1)
## Also, if we happen to have forgotten what was in the disp_time_binsBS object (as in
## which options where used to modify the original matrix) we can use the recall option:
summary(disp_time_binsBS, recall = TRUE)
## Note that the recall just prints text. The output of summary.dispRity is always a
## data.frame
summary_table <- summary(disp_time_binsBS, recall = TRUE)
class(summary_table)

## Finally we can see the results of the rarefaction analysis for the crown/stem data
head(summary(disp_crown_stemBS))
## This outputs a longer table with all the variations of the number of crown/stem taxa
## from the maximum (30 and 20) to the minimum (3). Note that the observed disparity
## is NA for rarefied data.
@

@@

The grey squares represent the confidence intervals (50 in dark grey and 90 in light grey) and the dot represents the mean.
However, it might be interesting to see how many taxa are present in each group.
This can be done via the \texttt{elements} option.
A third interesting option is to plot the rarefied data if a rarefaction was applied on the data with the \texttt{boot.matrix} function.
For displaying the rarefied data we can use the \texttt{rarefaction} option.
This option can be either \texttt{FALSE} (default) for always plotting the maximum number of taxa per series or \texttt{TRUE} to always plot the minimum.
Additionally, it can be a specific number to plot a specific number of taxa.
Let say we want both groups to have 20 taxa maximum.
We can see now that the confidence intervals overlap more than previously

<<plot1, fig.width=15, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 15, height = 5) ; par(mfrow = (c(1,3)), bty = "n")
## Plotting the score for both groups
plot(disp_crown_stemBS, ylim = c(25,40))
## Same one with the number of taxa for each data set
plot(disp_crown_stemBS, elements = TRUE, ylim = c(25,40))
## Same one with rarefied data (note how that affects the elements data as well!)
plot(disp_crown_stemBS, elements = TRUE, rarefaction = 20, ylim = c(25,40))
@

@@

<<plot2, fig.width=5, fig.height=8, out.width='0.5\\linewidth'>>=
## Graphical options
par(bty = "n")
## The rarefaction curves
plot(disp_crown_stemBS, elements = TRUE, rarefaction = "plot")
## Note that the function automatically draws the curves for each series and splits
## the plot screen accordingly.
@

@@

Additionally, for the \texttt{type = "discrete"} option, we might also decide to plot many boxplots (which can be a bit messy).
The \texttt{discrete\_type} option allows to switch between \texttt{"box"} or \texttt{"line"} for showing the results in one or the other format.
We can also plot the observed data when existing (i.e. not with some rarefaction levels) by using \texttt{observed = TRUE}.
Let's look at that with the time binned data (even though there is only three time bins).
Finally, we can use the \texttt{type = "continuous"} option that comes in handy for plotting continuous data like in the time sliced analysis (obviously...).

<<plot3, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
## The disparity-through-time data in a time-binned way (with lines rather than
## boxes) with the observed data as crosses
plot(disp_time_binsBS, discrete_type = "line", observed = TRUE)
## The disparity-through-time data in a time-sliced ways
plot(disp_time_slicesBS)

@@

<<plot5, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Graphical options
par(bty = "n")
## A plot with many options!
plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)), cent.tend = sd,
    type = "continuous", elements = "log", col = c("black", rainbow(10)),
    ylab = c("Disparity", "log(Diversity)"), xlab = "Time (in in units from
    past to present)", time.series = FALSE, observed = TRUE, main = "Many options...")
@
